---
title: "Airline Holiday Viz"
subtitle: "Passengers on flights departing Australia"
author: "Will Hore-Lacy"
date: "13/12/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

library(data.table)
library(sf)
library(leaflet)
library(ggplot2)
library(maps)
library(geosphere)
library(htmltools)

```

## Premise

In Australia the Christmas holidays coincide with the summer school holidays so it is often a time people travel. The Story Telling With Data challenge prompted to me considered if the airline travel patters were different in this holiday period. 

```{r downloadData, eval=FALSE}
# Download passenger data
# https://data.gov.au/dataset/international-airlines-traffic-by-city-pairs
download.file("https://data.gov.au/dataset/d9fbffaa-836f-4f52-80e8-324249ff269f/resource/ebcafd83-9514-4f72-a995-fe7ee90cb9da/download/city_pairs.csv",
              "./data/city_pairs.csv")

# Download airport data
# https://openflights.org/data.html
download.file("https://raw.githubusercontent.com/jpatokal/openflights/master/data/airports.dat",
              "./data/airports.csv")

```

```{r importData}

## Import data

city_pairs <- fread("./data/city_pairs.csv")


airportHeader <- c("Airport ID", "Name", "City",
                   "Country", "IATA", "ICAO",
                   "Latitude", "Longitude",
                   "Altitude", "Timezone", "DST", "Tz", 
                   "Type", "Source")

airports <- fread("./data/airports.csv", header = F)
names(airports) <- airportHeader


```

## Results

```{r totals}
# Make December and January the holiday period
city_pairs[Month_num %in% c(1,12), holiday:="yes"]
city_pairs[!(Month_num %in% c(1,12)), holiday:="no"]


# For mean of Australian ports
totalFlights <- dcast(city_pairs[Year>=2008],  holiday ~ .,
                     value.var = "Passengers_In", fun.aggregate=mean)


```

In general the number of flights increased in the holiday period from `r round(totalFlights[holiday=="no",.],0) ` to `r round(totalFlights[holiday=="yes",.],0)` (monthly average) so the results are skewed towards increases in passenger number, however there were some destinations the saw a reduction in the number of passengers.


```{r chart, fig.height=5.5, fig.width=8}


# For mean of Australian ports
aust_hol <- dcast(city_pairs[Year>=2008], ForeignPort + Country ~ holiday,
                     value.var = "Passengers_In", fun.aggregate=mean)

aust_hol[,delta := yes-no]
aust_hol[,absdelta:=abs(delta)]
aust_hol <- aust_hol[absdelta>0]

setorder(aust_hol,-delta)

len <- nrow(aust_hol)

aust_hol_top <- rbind(aust_hol[1:6],
                      aust_hol[(len-5):len])

aust_hol_top[, portNm:=paste0(ForeignPort,",\n",Country)]
aust_hol_top[, portNm:=factor(portNm,levels = aust_hol_top$portNm[1:12], ordered = T)]

aust_hol_top[delta>0,increase:="Increase"]
aust_hol_top[delta<0,increase:="Decrease"]


ggplot(aust_hol_top, aes(x=portNm, y=delta, fill=increase)) +
            geom_bar(stat="identity", position="dodge") +
            geom_hline(yintercept = 0, colour="grey", size=.4) +
            geom_vline(xintercept = 6.5, colour="grey", size=.4, linetype="dashed") +
            theme_light() +
            theme(axis.text.x = element_text(angle = 55, hjust=1, vjust=1),
                  panel.grid=element_blank(),
                  panel.border=element_blank(),
                  axis.line.y=element_line(colour="grey")) +
            labs(x="Destination", y="Change in Avg. monthly passengers", fill=NULL,
                 title="Change in airline passenger numbers for holiday period (Dec & Jan) ",
                 subtitle="Mean of all Australian Ports. Top and bottom 6 destinations.") +
            scale_fill_manual(values=c(Decrease="dark red",Increase="dark green")) + 
            guides(fill = guide_legend(reverse=T))

```


```{r, geocode}

# Add gps for origins and destinations

# Create dataset to use
pairs_hol <- dcast(city_pairs[Year>=2008], AustralianPort + ForeignPort + Country ~ holiday,
                     value.var = "Passengers_In", fun.aggregate=mean)

pairs_hol[,delta := yes-no]

pairs_hol[,absdelta:=abs(delta)]
pairs_hol <- pairs_hol[absdelta>0]


# Fix some city names
pairs_hol[grepl("Gold Coast", AustralianPort), AustralianPort:="Coolangatta"]
pairs_hol[grepl("Toowoomba Wellcamp", AustralianPort), AustralianPort:="Toowoomba"]

# Make field match city and country
pairs_hol[,origin:=paste0(AustralianPort,", Australia")]
pairs_hol[,destination:=paste0(ForeignPort,", ", Country)]
airports[,CityNm:=paste0(City,", ",Country)]

# Get gps based on city or aiport name
getGPS <- function(x) {
    
    # Look for exact match for name and country
    DT <- airports[CityNm==x]
    
    # Some heuristics to try and match other options
    if(nrow(DT)==0) {
        DT <- airports[Country==strsplit(x,", ")[[1]][2]]
        DT <-DT[agrep(strsplit(x,",")[[1]][1],Name, ignore.case = T, max.distance = .2)]
    }
    if(nrow(DT)==0) {
        DT <-airports[City==strsplit(x,",")[[1]][1]]
    }
    if(nrow(DT)==0) {
        DT <- airports[Country==strsplit(x,", ")[[1]][2]]
        DT <-DT[agrep(strsplit(x,",")[[1]][1],City, ignore.case = T, max.distance = .4)]
    }
    if(nrow(DT)==0) {
        DT <-airports[agrep(strsplit(x,",")[[1]][1],City, ignore.case = T, max.distance = .4)]
    }
    if(nrow(DT) >1) {
        DT2 <- DT[grepl("International",Name)]
        if(nrow(DT2)>0) {
            DT <- DT2
        }
    }
    if(nrow(DT)>1) {
        DT <- DT[1]
    }
    if(nrow(DT)==0) {
        return(data.table(Name=NA,Latitude=NA,Longitude=NA))
    }
    
    DT<-DT[,c("Name", "Latitude", "Longitude")]
    return(DT)
}

# get the gps for origin and destination
origin <- rbindlist(lapply(pairs_hol$origin,getGPS))
names(origin) <- paste0("origin_",names(origin))

destination <- rbindlist(lapply(pairs_hol$destination,getGPS))
names(destination) <- paste0("dest_",names(destination))

pairs_hol <- cbind(pairs_hol, origin, destination)


```

Change in the number of average monthly passengers in the December-January holiday period for origin destination pairs. Limited to the OD pairs with the top 15 increases and decreases.

```{r, fig.height=6, fig.width=10}

# Corrections allows for 'correct' plotting on world centered on Australia without wrapping
correctLon <- function(x) {

    if(x<(-50)) {
        x<-x+360
    }
    return(x)
}

# Converts two points into a great circle curved line 
makeLineSF <- function(x) {

    lon1 <- as.numeric(x["origin_Longitude"])
    lat1 <- as.numeric(x["origin_Latitude"])
    lon2 <- as.numeric(x["dest_Longitude"])
    lat2 <- as.numeric(x["dest_Latitude"])

    line1 <- gcIntermediate(c(lon1, lat1),
                            c(lon2, lat2),
                            n=100, addStartEnd=TRUE, breakAtDateLine = F)
    
    # remove 360 degrees from longitude
    line1[,1] <- sapply(line1[,1],correctLon)

    return(st_sf(id=1, geometry=st_sfc(st_linestring(line1, dim="XY"))))
    # return(makeGCline(lon1,lat1,lon2,lat2))

}

# Create lines to plot
pairs_hol <- pairs_hol[!is.na(dest_Latitude)]
pairs_hol_geom <-do.call("rbind", apply(pairs_hol,1,makeLineSF))

pairs_hol <- st_bind_cols(pairs_hol,pairs_hol_geom)

# Select top and bottom 15 locations
setorder(pairs_hol,-delta)
len <- nrow(pairs_hol)
pairs_hol_top30 <- rbind(pairs_hol[1:15,],
                      pairs_hol[(len-14):len,])

# Create map
pairsPal_top30 <- colorNumeric("RdYlGn", 
                       c(-max(pairs_hol_top30$absdelta),0,
                         max(pairs_hol_top30$absdelta)))

leaflet(pairs_hol_top30) %>%
    addProviderTiles("CartoDB.Positron") %>% #Provider ties
    addPolylines(weight = ~(absdelta/1000+2),color=~pairsPal_top30(delta),
                 label = ~paste0(origin," - ",destination,": ", round(delta,0)),
                 opacity=.7) %>%
    setView(lng = 144, lat = 10, zoom = 2) %>%
    addLegend(position = "bottomleft", pal=pairsPal_top30,
              values=c(-max(pairs_hol_top30$absdelta),0,
                         max(pairs_hol_top30$absdelta)),
              title = "Change in<br>Average Monthly<br>Passengers")
```


## Data

Two datasets are used for the visualisation:

* [Flight data](https://data.gov.au/dataset/international-airlines-traffic-by-city-pairs)
    * [city_pairs.csv](https://data.gov.au/dataset/d9fbffaa-836f-4f52-80e8-324249ff269f/resource/ebcafd83-9514-4f72-a995-fe7ee90cb9da/download/city_pairs.csv) (2008-2018 data only)
* [Airport data](https://openflights.org/data.html)
    * [airports.dat](https://raw.githubusercontent.com/jpatokal/openflights/master/data/airports.dat)



