---
title: "Airline Holiday Viz"
subtitle: "Passengers on flights departing Australia in the holiday period"
author: "Will Hore-Lacy"
date: "16/12/2018"
output:
  html_document:
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

library(data.table)
library(sf)
library(leaflet)
library(ggplot2)
library(maps)
library(geosphere)
library(htmltools)

```

## Premise

In Australia the Christmas holidays coincide with the summer school holidays so it is often a time people travel. The [Story Telling With Data challenge](https://www.storytellingwithdata.com/blog/2018/12/03/lets-visualize-the-holidays) prompted to me considered how airline travel patters were different in this holiday period. 

In the end I picked a problem that was too big for the offical challenge timeframe but here's what I came up with anyway... 

```{r downloadData, eval=FALSE}
# Download passenger data
# https://data.gov.au/dataset/international-airlines-traffic-by-city-pairs
download.file("https://data.gov.au/dataset/d9fbffaa-836f-4f52-80e8-324249ff269f/resource/ebcafd83-9514-4f72-a995-fe7ee90cb9da/download/city_pairs.csv",
              "./data/city_pairs.csv")

# Download airport data
# https://openflights.org/data.html
download.file("https://raw.githubusercontent.com/jpatokal/openflights/master/data/airports.dat",
              "./data/airports.csv")

```

```{r importData}

## Import data

city_pairs <- fread("./data/city_pairs.csv")


airportHeader <- c("Airport ID", "Name", "City",
                   "Country", "IATA", "ICAO",
                   "Latitude", "Longitude",
                   "Altitude", "Timezone", "DST", "Tz", 
                   "Type", "Source")

airports <- fread("./data/airports.csv", header = F)
names(airports) <- airportHeader


```

## Results

```{r totals}
# Make December and January the holiday period
city_pairs[Month_num %in% c(1,12), holiday:="yes"]
city_pairs[!(Month_num %in% c(1,12)), holiday:="no"]


# For mean of Australian ports
totalFlights <- dcast(city_pairs[Year>=2008],  holiday ~ .,
                     value.var = "Passengers_In", fun.aggregate=mean)


```

In general the number of flights increased in the holiday period from `r round(totalFlights[holiday=="no",.],0) ` to `r round(totalFlights[holiday=="yes",.],0)` passengers per monthly so the results are skewed towards increases in passenger number, however there were some destinations the saw a reduction in the number of passengers. Because the absolute changes in passenger numbers were not large, the visualisaitons use a the change in passenger numbers.


```{r chart, fig.height=5.8, fig.width=8}


# For mean of Australian ports
aust_hol <- dcast(city_pairs[Year>=2008], ForeignPort + Country ~ holiday,
                     value.var = "Passengers_In", fun.aggregate=mean)

aust_hol[,delta := yes-no]
aust_hol[,absdelta:=abs(delta)]
aust_hol <- aust_hol[absdelta>0]

setorder(aust_hol,-delta)

len <- nrow(aust_hol)

aust_hol_top <- rbind(aust_hol[1:6],
                      aust_hol[(len-5):len])

aust_hol_top[, portNm:=paste0(ForeignPort,",\n",Country)]
aust_hol_top[, portNm:=factor(portNm,levels = aust_hol_top$portNm[1:12], ordered = T)]

aust_hol_top[delta>0,increase:="Increase"]
aust_hol_top[delta<0,increase:="Decrease"]


ggplot(aust_hol_top, aes(x=portNm, y=delta, fill=increase)) +
            geom_bar(stat="identity", position="dodge") +
            geom_hline(yintercept = 0, colour="grey", size=.4) +
            geom_vline(xintercept = 6.5, colour="grey", size=.4, linetype="dashed") +
            geom_text(data=aust_hol_top[c(1,12)], aes(x=portNm, y=delta, label=round(delta,0)), 
                      colour="white", fontface = "bold", nudge_y=c(-150,155)) +
            theme_light() +
            theme(axis.text.x = element_text(angle = 55, hjust=1, vjust=1),
                  panel.grid=element_blank(),
                  panel.border=element_blank(),
                  axis.line.y=element_line(colour="grey")) +
            labs(x="Destination", y="Change in Avg. monthly passengers", fill=NULL,
                 title="Change in airline passenger numbers for holiday period (Dec & Jan) ",
                 subtitle="Mean of all Australian Ports 2008-2018 for top and bottom 6 destinations.") +
            scale_fill_manual(values=c(Decrease="dark red",Increase="dark green")) + 
            guides(fill = guide_legend(reverse=T)) +
            scale_y_continuous(breaks = c(-1000,0,1000,2000,3000))

```


```{r, geocode}

# Add gps for origins and destinations

# Create dataset to use
pairs_hol <- dcast(city_pairs[Year>=2008], AustralianPort + ForeignPort + Country ~ holiday,
                     value.var = "Passengers_In", fun.aggregate=mean)

# Calculate delta
pairs_hol[,delta := yes-no]

pairs_hol[,absdelta:=abs(delta)]
pairs_hol <- pairs_hol[absdelta>0]

# Delta as a percentage of non-holiday period
pairs_hol[,deltaPer := delta/no]

# Fix some city names
pairs_hol[grepl("Gold Coast", AustralianPort), AustralianPort:="Coolangatta"]
pairs_hol[grepl("Toowoomba Wellcamp", AustralianPort), AustralianPort:="Toowoomba"]

# Make field match city and country
pairs_hol[,origin:=paste0(AustralianPort,", Australia")]
pairs_hol[,destination:=paste0(ForeignPort,", ", Country)]
airports[,CityNm:=paste0(City,", ",Country)]

# Get gps based on city or aiport name
getGPS <- function(x) {
    
    # Look for exact match for name and country
    DT <- airports[CityNm==x]
    
    # Some heuristics to try and match other options
    if(nrow(DT)==0) {
        DT <- airports[Country==strsplit(x,", ")[[1]][2]]
        DT <-DT[agrep(strsplit(x,",")[[1]][1],Name, ignore.case = T, max.distance = .2)]
    }
    if(nrow(DT)==0) {
        DT <-airports[City==strsplit(x,",")[[1]][1]]
    }
    if(nrow(DT)==0) {
        DT <- airports[Country==strsplit(x,", ")[[1]][2]]
        DT <-DT[agrep(strsplit(x,",")[[1]][1],City, ignore.case = T, max.distance = .4)]
    }
    if(nrow(DT)==0) {
        DT <-airports[agrep(strsplit(x,",")[[1]][1],City, ignore.case = T, max.distance = .4)]
    }
    if(nrow(DT) >1) {
        DT2 <- DT[grepl("International",Name)]
        if(nrow(DT2)>0) {
            DT <- DT2
        }
    }
    if(nrow(DT)>1) {
        DT <- DT[1]
    }
    if(nrow(DT)==0) {
        return(data.table(Name=NA,Latitude=NA,Longitude=NA))
    }
    
    DT<-DT[,c("Name", "Latitude", "Longitude")]
    return(DT)
}

# get the gps for origin and destination
origin <- rbindlist(lapply(pairs_hol$origin,getGPS))
names(origin) <- paste0("origin_",names(origin))

destination <- rbindlist(lapply(pairs_hol$destination,getGPS))
names(destination) <- paste0("dest_",names(destination))

pairs_hol <- cbind(pairs_hol, origin, destination)


```

The same plot can be generated individually for each Australian port, however a map below provides some extra information. The map shows the change in average monthly passengers in the December-January holiday period for origin destination pairs. Limited to the OD pairs with the top 12 increases and decreases.

```{r, fig.height=6, fig.width=10}

# Corrections allows for 'correct' plotting on world centered on Australia without wrapping
correctLon <- function(x) {

    if(x<(-50)) {
        x<-x+360
    }
    return(x)
}

# Converts two points into a great circle curved line 
makeLineSF <- function(x) {

    lon1 <- as.numeric(x["origin_Longitude"])
    lat1 <- as.numeric(x["origin_Latitude"])
    lon2 <- as.numeric(x["dest_Longitude"])
    lat2 <- as.numeric(x["dest_Latitude"])

    line1 <- gcIntermediate(c(lon1, lat1),
                            c(lon2, lat2),
                            n=120, addStartEnd=TRUE, breakAtDateLine = F)
    
    # remove 360 degrees from longitude
    line1[,1] <- sapply(line1[,1],correctLon)

    return(st_sf(id=1, geometry=st_sfc(st_linestring(line1, dim="XY"))))
    # return(makeGCline(lon1,lat1,lon2,lat2))

}

# Create lines to plot
pairs_hol <- pairs_hol[!is.na(dest_Latitude)]
pairs_hol_geom <-do.call("rbind", apply(pairs_hol,1,makeLineSF))

pairs_hol <- st_bind_cols(pairs_hol,pairs_hol_geom)
# pairs_hol <- cbind(pairs_hol,pairs_hol_geom)

# Select top and bottom 15 locations
setorder(pairs_hol,-delta)
len <- nrow(pairs_hol)
pairs_hol_top24 <- rbind(pairs_hol[1:12,],
                      pairs_hol[(len-11):len,])

title_txt <- '<div style="text-align:right"><strong>Change in Avg. monthly passenger numbers</strong><br>Top 12 increases and decreases</div>'

# Create map
# Pal. Reverse to get green at the top of the legend
pairsPal_top24 <- colorNumeric("RdYlGn", 
                       c(-max(pairs_hol_top24$absdelta),0,
                         max(pairs_hol_top24$absdelta)),
                       reverse = TRUE)

leaflet(pairs_hol_top24) %>%
    # addProviderTiles("Stamen.Toner") %>% #Provider ties
    addProviderTiles("CartoDB.DarkMatterNoLabels", options = providerTileOptions(opacity=.78)) %>%
    addProviderTiles("CartoDB.DarkMatterOnlyLabels", options = providerTileOptions(opacity=.35)) %>% #Provider ties
    addPolylines(data = pairs_hol_top24[c(1,nrow(pairs_hol_top24)),], 
                 weight = ~(absdelta/1000+4.7),color="blue", # -delta to reverse values
                 opacity=.80) %>%
    addPolylines(weight = ~(absdelta/1000+2),color=~(pairsPal_top24(-delta)), # -delta to reverse values
                 label = ~paste0(origin," - ",destination,": ", round(delta,0)),
                 opacity=.95) %>%
    setView(lng = 144, lat = 10, zoom = 2) %>%
    addControl(title_txt, position = "topright") %>%
    addLegend(position = "bottomleft", pal=pairsPal_top24,
              values=c(-max(pairs_hol_top24$absdelta),0,
                         max(pairs_hol_top24$absdelta)),
              title = "Change in<br>Avg. Monthly<br>Passengers",
              labFormat = labelFormat(transform = function(x) sort(x, decreasing = TRUE))) # reverse legend values
```
<br>

The first interesting patter is that it apperars that the data only show the first port visited rather than final destination, i.e. no locaitons in Europe but flights to major inerchange hubs. The second is that it shows a few differences in the destinations with large changes. 

The top increased and decrease routes are highlighted in blue:

* `r pairs_hol_top24$AustralianPort[1]` to `r pairs_hol_top24$ForeignPort[1]`: `r as.integer(pairs_hol_top24$delta[1])`
* `r pairs_hol_top24$AustralianPort[nrow(pairs_hol_top24)]` to `r pairs_hol_top24$ForeignPort[nrow(pairs_hol_top24)]`: `r round(pairs_hol_top24$delta[nrow(pairs_hol_top24)], 0)`


## Data

Two datasets are used for the visualisation:

* [Flight data](https://data.gov.au/dataset/international-airlines-traffic-by-city-pairs)
    * [city_pairs.csv](https://data.gov.au/dataset/d9fbffaa-836f-4f52-80e8-324249ff269f/resource/ebcafd83-9514-4f72-a995-fe7ee90cb9da/download/city_pairs.csv) (2008-2018 data only)
* [Airport data](https://openflights.org/data.html)
    * [airports.dat](https://raw.githubusercontent.com/jpatokal/openflights/master/data/airports.dat)



