---
title: "Airline Viz EDA"
author: "Will Hore-Lacy"
date: "06/12/2018"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(sf)
library(leaflet)
library(ggplot2)
library(maps)
library(geosphere)

```

## Import data

```{r}

city_pairs <- fread("./data/city_pairs.csv")
aust_flights <- fread("./data/international_airline_activity_opfltsseats.csv")
seats <- fread("./data/seat_utilisation.csv")

head(city_pairs,1)

```


Aggregate data as a monthly average from 2008

```{r}

pairs_month <- dcast(city_pairs[Year>=2008], AustralianPort + ForeignPort + Country + Month_num ~.,
                     value.var = "Passengers_In", fun.aggregate=mean)

setnames(pairs_month,".","Passengers_In")

city_pairs[Month_num %in% c(1,12), holiday:="yes"]
city_pairs[!(Month_num %in% c(1,12)), holiday:="no"]

pairs_month[Month_num %in% c(1,12), holiday:="yes"]
pairs_month[!(Month_num %in% c(1,12)), holiday:="no"]

pairs_hol <- dcast(city_pairs[Year>=2008], AustralianPort + ForeignPort + Country ~ holiday,
                     value.var = "Passengers_In", fun.aggregate=mean)

pairs_hol[,delta := yes-no]

```


## Graphs

For each origin (facet), plot top 10 destinations vs passengers, by holidays vs not (fill).

```{r, fig.width=6, fig.height=5}

top_ports <- dcast(city_pairs, AustralianPort ~ ., value.var="Passengers_In", fun.aggregate=sum)
setorder(top_ports,-.)
top_ports <- top_ports[1:7]

pairs_hol[,absdelta:=abs(delta)]
pairs_hol <- pairs_hol[absdelta>0]

pairs_hol_top <- data.table()

# for(dest in top_ports$AustralianPort) {
#     
#     DT <- pairs_hol[AustralianPort==dest]
#     setorder(DT,-absdelta)
#     pairs_hol_top <- rbind(pairs_hol_top,
#                            DT[1:10])
#     
# }

for(origin in top_ports$AustralianPort) {
    
    DT <- pairs_hol[AustralianPort==origin]
    setorder(DT,-absdelta)
    DT_melt <- melt(DT[1:12, 1:5], id.vars = c("AustralianPort","ForeignPort","Country"),
                       variable.name = "holiday", value.name = "Passengers_In")
    
    g <- ggplot(DT_melt, aes(x=paste0(ForeignPort,",\n",Country), y=Passengers_In, fill=holiday)) +
            geom_bar(stat="identity", position="dodge") +
            facet_wrap("AustralianPort") +
            theme_bw() +
            theme(axis.text.x = element_text(angle = 55, hjust=1, vjust=1)) +
            labs(x="Destination", y="Avg. monthly passengers", fill="Holiday\nperiod") 
    
    print(g)
}

# pairs_hol_melt <- melt(pairs_hol_top[,1:5], id.vars = c("AustralianPort","ForeignPort","Country"),
#                        variable.name = "holiday", value.name = "Passengers_In")

# ggplot(pairs_hol_melt, aes(x=ForeignPort, y=Passengers_In, fill=holiday)) +
#     geom_bar(stat="identity", position="dodge") +
#     facet_wrap("AustralianPort") +
#     theme(axis.text.x = element_text(angle = 90))



```


Remove the origin, sum across the country, same plot but set-up for top and bottom 6.

Also difference plot

```{r, fig.height=5.5, fig.width=8}

# For mean of Australian ports
aust_hol <- dcast(city_pairs[Year>=2008], ForeignPort + Country ~ holiday,
                     value.var = "Passengers_In", fun.aggregate=mean)

#### For sum of Australian ports

# aust_hol <- dcast(pairs_month, ForeignPort + Country + holiday + Month_num ~ .,
#                      value.var = "Passengers_In", fun.aggregate=sum)
# 
# 
# setnames(aust_hol,".","Passengers_In")
# 
# aust_hol <- dcast(aust_hol, ForeignPort + Country ~ holiday,
#                      value.var = "Passengers_In", fun.aggregate=mean)
####

aust_hol[,delta := yes-no]
aust_hol[,absdelta:=abs(delta)]
aust_hol <- aust_hol[absdelta>0]

setorder(aust_hol,-delta)

len <- nrow(aust_hol)

aust_hol_top <- rbind(aust_hol[1:6],
                      aust_hol[(len-5):len])

aust_hol_top[, portNm:=paste0(ForeignPort,",\n",Country)]
aust_hol_top[, portNm:=factor(portNm,levels = aust_hol_top$portNm[1:12], ordered = T)]

aust_hol_top[delta>0,increase:="Increase"]
aust_hol_top[delta<0,increase:="Decrease"]


ggplot(aust_hol_top, aes(x=portNm, y=delta, fill=increase)) +
            geom_bar(stat="identity", position="dodge") +
            geom_hline(yintercept = 0, colour="grey", size=.4) +
            geom_vline(xintercept = 6.5, colour="grey", size=.4, linetype="dashed") +
            theme_light() +
            theme(axis.text.x = element_text(angle = 55, hjust=1, vjust=1),
                  panel.grid=element_blank(),
                  panel.border=element_blank(),
                  axis.line.y=element_line(colour="grey")) +
            labs(x="Destination", y="Change in Avg. monthly passengers", fill=NULL,
                 title="Change in airline passenger numbers for holiday period (Dec & Jan) ",
                 subtitle="Mean of all Australian Ports. Top and bottom 6 destinations.") +
            scale_fill_manual(values=c(Decrease="dark red",Increase="dark green")) + 
            guides(fill = guide_legend(reverse=T))

aust_hol_top_melt <- melt(aust_hol_top[,c(1:4,7)], id.vars = c("ForeignPort","Country", "portNm"),
                       variable.name = "holiday", value.name = "Passengers_In")

# aust_hol_top_melt[, portNm:=paste0(ForeignPort,",\n",Country)]
# aust_hol_top_melt[, portNm:=factor(portNm,levels = aust_hol_top_melt$portNm[1:12], ordered = T)]

ggplot(aust_hol_top_melt, aes(x=portNm, y=Passengers_In, fill=holiday)) +
            geom_bar(stat="identity", position="dodge") +
            theme_bw() +
            theme(axis.text.x = element_text(angle = 55, hjust=1, vjust=1)) +
            labs(x="Destination", y="Avg. monthly passengers", fill="Holiday\nperiod") 

```

## Map Viz

Departing Australian cities. Line thickness for volume to each destination.

1. First get lat and lon for each city
2. Get flight arc info
3. Plot
    1. Leaflet
    2. ggmap

```{r, geocode}

# Import airport data
airportHeader <- c("Airport ID", "Name", "City",
                   "Country", "IATA", "ICAO",
                   "Latitude", "Longitude",
                   "Altitude", "Timezone", "DST", "Tz", 
                   "Type", "Source")

# https://openflights.org/data.html
#https://raw.githubusercontent.com/jpatokal/openflights/master/data/airports.dat
airports <- fread("./data/airports.csv", header = F)
names(airports) <- airportHeader


# airports[grepl("Toowoomba",City)]
# airports[grepl("Toowoomba Wellcamp",City)]


# Fix some city names
pairs_hol[grepl("Gold Coast", AustralianPort), AustralianPort:="Coolangatta"]
pairs_hol[grepl("Toowoomba Wellcamp", AustralianPort), AustralianPort:="Toowoomba"]

# Make field match city and country
pairs_hol[,origin:=paste0(AustralianPort,", Australia")]
pairs_hol[,destination:=paste0(ForeignPort,", ", Country)]
airports[,CityNm:=paste0(City,", ",Country)]

# Get gps based on city or aiport name
getGPS <- function(x) {
    
    DT <- airports[CityNm==x]
    
    if(nrow(DT)==0) {
        DT <- airports[Country==strsplit(x,", ")[[1]][2]]
        DT <-DT[agrep(strsplit(x,",")[[1]][1],Name, ignore.case = T, max.distance = .2)]
    }
    if(nrow(DT)==0) {
        DT <-airports[City==strsplit(x,",")[[1]][1]]
    }
    if(nrow(DT)==0) {
        DT <- airports[Country==strsplit(x,", ")[[1]][2]]
        DT <-DT[agrep(strsplit(x,",")[[1]][1],City, ignore.case = T, max.distance = .4)]
    }
    if(nrow(DT)==0) {
        DT <-airports[agrep(strsplit(x,",")[[1]][1],City, ignore.case = T, max.distance = .4)]
    }
    if(nrow(DT) >1) {
        DT2 <- DT[grepl("International",Name)]
        if(nrow(DT2)>0) {
            DT <- DT2
        }
    }
    if(nrow(DT)>1) {
        DT <- DT[1]
    }
    if(nrow(DT)==0) {
        return(data.table(Name=NA,Latitude=NA,Longitude=NA))
    }
    
    DT<-DT[,c("Name", "Latitude", "Longitude")]
    return(DT)
}

# getGPS("Toowoomba, Australia")
# debugonce(getGPS)

# cbind(unique(city_pairs$origin),
#       rbindlist(lapply(unique(city_pairs$origin),getGPS))
# )

# get the gps for origina and destination
origin <- rbindlist(lapply(pairs_hol$origin,getGPS))
names(origin) <- paste0("origin_",names(origin))
destination <- rbindlist(lapply(pairs_hol$destination,getGPS))
names(destination) <- paste0("dest_",names(destination))

pairs_hol <- cbind(pairs_hol, origin, destination)

# sum(is.na(destination))



```

```{r, fig.height=6, fig.width=10}

# pairs_hol_backup <- pairs_hol
# pairs_hol <- pairs_hol_backup
# pairs_hol$origin_Longitude[1]

# makeGCline <- function(lon1, lat1, lon2, lat2) {
#     line1 <- gcIntermediate(c(lon1, lat1), 
#                             c(lon2, lat2), 
#                             n=50, addStartEnd=TRUE)
#     return(st_linestring(line1, dim="XY")) # or something similar
# }

correctLon <- function(x) {
    # if(x>180) {
    #     x<-x-360
    # }
    if(x<(-50)) {
        x<-x+360
    }
    return(x)
}

makeLineSF <- function(x) {

    lon1 <- as.numeric(x["origin_Longitude"])
    lat1 <- as.numeric(x["origin_Latitude"])
    lon2 <- as.numeric(x["dest_Longitude"])
    lat2 <- as.numeric(x["dest_Latitude"])

    line1 <- gcIntermediate(c(lon1, lat1),
                            c(lon2, lat2),
                            n=100, addStartEnd=TRUE, breakAtDateLine = F)
    
    # remove 360 degrees from longitude
    line1[,1] <- sapply(line1[,1],correctLon)

    return(st_sf(id=1, geometry=st_sfc(st_linestring(line1, dim="XY"))))
    # return(makeGCline(lon1,lat1,lon2,lat2))

}

pairs_hol <- pairs_hol[!is.na(dest_Latitude)]
pairs_hol_geom <-do.call("rbind", apply(pairs_hol,1,makeLineSF))

pairs_hol <- st_bind_cols(pairs_hol,pairs_hol_geom)

# st_sf(1:2, geometry=st_sfc(apply(pairs_hol[1:2],1,makeLineSF)))
# 
# df <- st_sf(id = 1:nrows, geometry = st_sfc(lapply(1:nrows, function(x) st_geometrycollection())))
# 
# apply(pairs_hol[1],1,makeLineSF)

# 
# 
# a <- do.call("st_line",apply(pairs_hol[Country=="USA"][1],1,makeLineSF))
# a <- apply(pairs_hol[Country=="USA"][1],1,makeLineSF)
# st_coordinates(a[[1]])
# makeLineSF(pairs_hol[1])

# makeLineSF(pairs_hol[Country=="USA"][1])
# debugonce(makeLineSF)


# paris_hol_geom <- c()
# 
# for(i in 1:nrow(pairs_hol)) {
#     
#     paris_hol_geom <- c(paris_hol_geom,
#                             makeGCline(pairs_hol[i,origin_Longitude],
#                                      pairs_hol[i,origin_Latitude],
#                                      pairs_hol[i,dest_Longitude],
#                                      pairs_hol[i,dest_Latitude])
#                         )
#     
#     
#     
#     
# }




# c(pairs_hol$origin_Longitude[1], pairs_hol$origin_Latitude[1]), 
#                             c(pairs_hol$dest_Longitude[1], pairs_hol$dest_Latitude[1])
# inter_sf <- st_as_sf(as.data.frame(inter), coords = c(1,2))
# plot(inter_sf$geometry)
# 
# inter_sf_ln <- st_linestring(inter, dim="XY")
# plot(inter_sf_ln)

# map("world")
# lines(pairs_hol$geometry,col="red", cex=5)

# plot(pairs_hol$geometry)

holiday_lines <- pairs_hol[order(-pairs_hol$yes),][1:30,]

nonholiday_lines <- pairs_hol[order(-pairs_hol$no),][1:30,]

library(RColorBrewer)
holPal <- colorNumeric("OrRd", 
                       c(min(min(nonholiday_lines$no),min(holiday_lines$yes))-10000,
                         max(max(nonholiday_lines$no),max(holiday_lines$yes))))

leaflet(holiday_lines) %>%
    addProviderTiles("CartoDB.Positron") %>% #Provider ties
    addPolylines(weight=~(yes/5000+.5), color=~holPal(yes)) %>%
    setView(lng = 144, lat = 25, zoom = 2)

leaflet(nonholiday_lines) %>%
    addProviderTiles("CartoDB.Positron") %>% #Provider ties
    addPolylines(weight=~(no/5000+.5), color=~holPal(no)) %>%
    setView(lng = 144, lat = 25, zoom = 2)


# plot(nonholiday_lines$geometry)


# library(googleway)
# 
# set_key("AppKey")
# 
# google_map() %>%
#     add_polylines(data = holiday_lines)
# 

```

```{r, fig.width=10, fig.height=8}
wrld <- borders("world", colour="gray50", fill="gray50",xlim = c(-50, 310))
world2 <- map_data('world', wrap=c(-30,330), ylim=c(-50,90))

# plot the result
# ggplot(holiday_lines) + wrld +
#     geom_sf() +
#     coord_sf(xlim = c(-50, 310), ylim = c(-50, 75))

ggplot(holiday_lines) + 
    geom_polygon(data=world2, aes(x = long, y = lat, group = group), col = "white", fill="grey")+
    geom_sf(aes(colour=yes, size=yes)) +
    theme_bw() +
    scale_colour_continuous(low="orange", high="red") +
    scale_size_continuous(range=c(.02,4),
                          limits=c(min(min(nonholiday_lines$no),min(holiday_lines$yes)),
                         max(max(nonholiday_lines$no),max(holiday_lines$yes)))) +
    labs(title="Holiday period", x=NULL, y=NULL) +
    theme(legend.position="bottom")

ggplot(nonholiday_lines) + 
    geom_polygon(data=world2, aes(x = long, y = lat, group = group), col = "white", fill="grey")+
    geom_sf(aes(colour=no, size=no)) +
    theme_bw() +
    scale_colour_continuous(low="orange", high="red") +
    scale_size_continuous(range=c(.02,4),
                          limits=c(min(min(nonholiday_lines$no),min(holiday_lines$yes)),
                         max(max(nonholiday_lines$no),max(holiday_lines$yes)))) +
    labs(title="Non-holiday period", x=NULL, y=NULL) +
    theme(legend.position="bottom")



```

```{r, fig.width=10, fig.height=6}

####  DO A PLOT WITH THE CHANGE LINES  ####

setorder(pairs_hol,-delta)
len <- nrow(pairs_hol)
pairs_hol_top30 <- rbind(pairs_hol[1:15,],
                      pairs_hol[(len-14):len,])




pairsPal_top30 <- colorNumeric("RdYlGn", 
                       c(-max(pairs_hol_top30$absdelta),0,
                         max(pairs_hol_top30$absdelta)))

library(htmltools)

leaflet(pairs_hol_top30) %>%
    addProviderTiles("CartoDB.Positron") %>% #Provider ties
    addPolylines(weight = ~(absdelta/1000+2),color=~pairsPal_top30(delta),
                 label = ~paste0(origin," - ",destination,": ", round(delta,0)),
                 opacity=.7) %>%
    setView(lng = 144, lat = 10, zoom = 2) %>%
    addLegend(position = "bottomleft", pal=pairsPal_top30,
              values=c(-max(pairs_hol_top30$absdelta),0,
                         max(pairs_hol_top30$absdelta)))



```


```{r, eval=FALSE}

st_coordinates(pairs_hol[pairs_hol$Country=="USA", "geometry"])
# unique(pairs_hol$Country)


berlin <- c(52.51, 13.4)
losangeles <- c(34.05, -118.24)
santiago <- c(-33.44, -70.71)
tokio <- c(35.69, 139.69)
sydney <- c(-33.91, 151.08)
capetown <- c(-33.91, 18.41)
calgary <- c(51.05, -114.08)
hammerfest <- c(70.67, 23.68)
barrow <- c(71.29, -156.76)
Adelaide <- c(-34.945000000, 138.53100)
USA <- c(21.318701000, -157.92200)
# df <- as.data.frame(rbind(hammerfest, calgary, losangeles, santiago, capetown, tokio, barrow))
df<-as.data.frame(rbind(Adelaide, USA))
names(df) <- c("lat","lng")

leaflet(df) %>%
addProviderTiles(providers$CartoDB.Positron) %>%
addGeodesicPolylines(lng = ~lng, lat = ~lat, weight = 5, color = "red",
steps = 50, opacity = 1) #%>%
# addCircleMarkers(df, lat = ~lat,lng = ~lng, radius = 3, stroke = FALSE,
# fillColor = "black", fillOpacity = 1)
# 

```


